Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model

### 2022-09-16 Rmarkdown/paper.Rmd
pre-submission edits (revision 2)

Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting. They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995]. The most common models for pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].

@@

The primary goal of a pipe loop study is to compare two or more corrosion control treatments [@wysock_statistical_1995]. But the data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given their fairly rigid assumptions. Often, a more flexible approach is required, and Bayesian methods are inherently flexible since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].

@@

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a polytetrafluoroethylene (PTFE) filter and a vacuum filter apparatus. Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high. In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters [@minning_systematic_2015].

@@

The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a). And while the model assumes that seasonality strength does not vary over time, some previous work has suggested that orthophosphate may damp seasonal variation in lead [@colling_measurement_1987]. This effect could be incorporated by allowing the seasonal smoother to vary with orthophosphate dose.

@@

$$
(5)~~
\begin{align}
likelihood: \\
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu)~~[observed] \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu)~~[left \text{-} censored] \\
y_t | censored_t = 2 \sim g(\mu_t,\sigma,\nu)~~[interval \text{-} censored] \\
\\
model~for~\mu_t: \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
priors: \\
\sigma \sim Half \text{-} T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$. ($Gamma$ and $N$ represent the gamma and Gaussian distributions.)

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
\alpha \sim T(0.1, 2.5, 3)~~[prior~on~global~intercept] \\
\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3)~~[prior~on~\sigma~of~pipe~averages] \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:

$$
(7)~~
\begin{align}
generalized~additive~model~(GAM): \\
f_j(t) = X_j\beta_j + Z_jb_j \\
\\
priors~on~GAM~coefficients: \\
\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
b_j \sim N(0, \sigma_b)~~[penalized] \\
\sigma_b \sim Half \text{-} N(0, 0.5) (seasonal~smoother) \\
\sigma_b \sim  Half \text{-} T(0, 2.5, 3) (global/group \text{-}level~smoothers) \\ 
\end{align}
$$

@@

In a typical pipe loop study, effluent lead concentrations decrease sharply after commissioning. Assuming inputs to the system remain constant, a period of relative stability usually follows [@wysock_statistical_1995]. This pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured directly by the model (see *Trends in particulate lead release*).

@@

Our approach can be modified to suit the needs of diverse analyses. While time is the sole independent variable here, water quality parameters can easily be added as predictors if inferences concerning their effects are of interest. Moreover, any parameter in the likelihood can be modeled---including the standard deviation if model residuals are heteroskedastic. The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical]. But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time.

>>>>

Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting model parameters. Nonlinearity can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995]. The most common models for pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].

@@

The primary goal of a pipe loop study is to compare two or more corrosion control treatments [@wysock_statistical_1995]. But the data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given the necessary assumptions. Often, a more flexible approach is required, and Bayesian methods are inherently flexible since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].

@@

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a polytetrafluoroethylene (PTFE) filter and a vacuum filter apparatus. Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high. In hard waters such as the feedwater for the model system, though, this bias is likely to be much lower than in soft waters [@minning_systematic_2015].

@@

The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a). And while the model assumes that seasonality strength does not vary over time, some previous work has suggested that orthophosphate may damp seasonal variation in lead [@colling_measurement_1987]. This effect could be incorporated by allowing the seasonal smoother to vary with the orthophosphate dose.

@@

$$
(5)~~
\begin{align}
\text{likelihood}: \\
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu)~~[observed] \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu)~~[left \text{-} censored] \\
y_t | censored_t = 2 \sim g(\mu_t,\sigma,\nu)~~[interval \text{-} censored] \\
\\
\text{model for }\mu_t\text{:} \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s) \\
\\
\text{priors}: \\
\sigma \sim Half \text{-} T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$. ($Gamma$ and $N$ represent the gamma and Gaussian distributions, parameterized by the mean and shape parameter/standard deviation, respectively.)

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
\alpha \sim T(0.1, 2.5, 3)~~[prior~on~global~intercept] \\
\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3)~~[prior~on~\sigma~of~pipe~averages] \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$. The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:

$$
(7)~~
\begin{align}
\text{generalized additive model (GAM)}: \\
f_j(t) = X_j\beta_j + Z_jb_j \\
\\
\text{priors on GAM parameters}: \\
\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
b_j \sim N(0, \sigma_b)~~[penalized] \\
\sigma_b \sim Half \text{-} N(0, 0.5) [seasonal~smoother] \\
\sigma_b \sim  Half \text{-} T(0, 2.5, 3) [global/group \text{-}level~smoothers] \\ 
\end{align}
$$

@@

In a typical pipe loop study, effluent lead concentrations decrease sharply after commissioning. Assuming inputs to the system remain constant, a period of relative stability usually follows [@wysock_statistical_1995]. This pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized (Figure 8a). The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured directly by the model (see *Trends in particulate lead release*).

@@

Our approach can be modified to suit the needs of diverse analyses. While time is the sole independent variable here, water quality parameters can easily be added as predictors if inferences concerning their effects are of interest. Moreover, any parameter in the likelihood can be modeled---including the standard deviation if model residuals are heteroskedastic. The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical]. But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real time.
